\vspace{5cm} \normalsize \textsc{\NetIDb} \\ [2.0cm] \HRule{0.5pt} \\ \LARGE \textbf{\uppercase{\documenttitle}} \HRule{2pt} \\ [0.5cm] \normalsize \textsc{Sergio Castañeiras Morales\\NIU:1598456\\ }\normalsize \vspace*{5\baselineskip}

### 2025-05-01 Manuscript/Plantilla memòria TFG 2024_25.tex
writting

%-------------LINKS AND LOOKS-----------%
\usepackage{fullpage}
\usepackage{afterpage}
\usepackage{lastpage}
\usepackage{fancyhdr}
\usepackage{forest}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage{hyperref}
\usepackage{pgfplots}
\usepackage{xurl}
\usepackage{pdflscape}
%-----------------FIGURES AND TABLES----------------%
\usepackage{array}
%\usepackage{wrapfig}
\usepackage{caption}
\DeclareCaptionType{scheme}
%\usepackage{subcaption}
\usepackage{tabularx, booktabs, makecell, multicol, multirow, bigstrut} %for excel2latex
\usepackage{float}

@@

Determining an experimental \gls{ic50} value can be both costly and time-consuming\footnote{Usually this parameter is computed throughout the Cheng Prusoff Equation (Def. (\ref{definitionChengPrusoff})) with experimental data.}. On the other hand, \gls{ai} provides an alternative by offering highly accurate predictions based on existing data, optimising research processes, and accelerating scientific discovery. In this scenario this project aims to implement artificial intelligence in computational chemistry, concretely, using \gls{ai}-based algorithms to predict a drug’s inhibition potential\cite{BookIC50} for a given protein with a relativelly good accuracy \footnote{The accuracy of the ML models is discussed on the Results and Discussion section (\ref{sectionResultsAndDiscussion}).}. To achieve this, we make use of the ChEMBL database\cite{ChemblDatabase}, a vast repository of bioactive molecules with drug-like properties. We extract all known molecular data with a documented inhibition potential for the target protein, creating a comprehensive dataset. The chemical descriptors of each molecule in the database are then computed using AlvaDesk\cite{MauriMolecularDescriptorsBook}\cite{AlvaDescSecondPaper} software. Around 5000 descriptors are calculated\cite{DescriptorsBook}, which comprehend from the elemental molecular weight to the complex equipotential electronic surface, providing critical information about each compound’s behaviour. The resulting dataset is subsequently used to train \gls{ai} models, enabling them to predict the inhibition potential of unknown compounds. Finally, we evaluate the reliability of each model by testing it against real experimental data.

@@

\begin{figure}[H]
	\centering
	\includegraphics[width = \textwidth, trim={0.85cm 0.3cm 1.5cm 1.35cm}, clip]{../Plots/DataGapRepresentationIC50percentageEresed15.0Gap_min_32Gap_max_53.pdf}
	\caption{\gls{ic50} values from our dataset and generation of a 20\% gap.}
	\label{figureGraphicRecurrecyIC50}
\end{figure}

@@

\begin{figure}[H]
\begin{tikzpicture}
\begin{axis}[
    width=0.45\textwidth,
    height=8cm,
    xlabel={$\%_{\text{Erased}}$},
    ylabel={$r^2$},
    xmin=0, xmax=100,
    ymin=0.94, ymax=1.0,
    xtick={0, 10, 20, 30, 40, 50, 60, 70, 80, 90, 100},
    ytick={0.9, 0.92, 0.94, 0.96, 0.98, 1.0},
    legend pos=south east,
    legend style={font=\small},
    tick label style={font=\small},
    label style={font=\small},
]
\addplot[
    color=magenta,
    mark=diamond*,
    thick
] table[x index=0, y index=1, col sep=space] {GeneralSources/ErasedPercentageVSRSquare.txt};
\end{axis}
\end{tikzpicture}
\begin{tikzpicture}
\begin{axis}[
    width=0.45\textwidth,
    height=8cm,
    xlabel={$\%_{\text{Erased}}$},
	ylabel={Mean Squared Error (nM$^2$)},
    xmin=0, xmax=100,
    ymin=0, ymax=200,
    xtick={0, 10, 20, 30, 40, 50, 60, 70, 80, 90, 100},
    legend pos=north east,
    legend style={font=\small},
    tick label style={font=\small},
    label style={font=\small}
]
\addplot[
    color=magenta,
    mark=diamond*,
    thick
] table[x index=0, y index=1, col sep=space] {GeneralSources/ErasedPercentageVSMeanSquaredError.txt};
\end{axis}
\end{tikzpicture}
\caption{Plots of $\%_{\text{Erased}}$ vs $r^2$ and $\%_{\text{Erased}}$ vs Mean Squared Error. For a maximum $IC_{50}$ = 200 nM and a minimum of 0 nM,  taking into account all the chemical descriptors, the 20.0\% of the remaining data is saved for testing and the  80.0\% for training. The training is done with 200 trees.}
\label{FigPlotErasedPercentageVSRSquare}
\end{figure}

@@

Furthermore, we may also show the statistics and reliability of this results in Table (\ref{TableStatisitcsMainResults}).

\begin{table}[H]
    \centering	
	\caption{Statistics for the computation of the results from tables (\ref{TableMainResults1}) and (\ref{TableMainResults2}).}
	\label{TableStatisitcsMainResults}
	\begin{tabular}{c|c}
		Mean Squared Error       & 1202.70\\\hline
		$r^2$ 	                 & 0.90\\\hline
		ClassificationAcuracy    & 0.94\\\hline
		MatthewsCorrelationFactor& 0.87
    \end{tabular}
\end{table}

In Figure (\ref{figureGraphicGbindingIC50}) we can see a clear correlation between our initial intuition. Lower values of free eneregy (the most negative ones) correspond to low \gls{ic50} values i.e. highly active molecules

@@

\begin{figure}[H]
	\centering
	\includegraphics[width = \textwidth, trim={0cm 0cm 0cm 0cm}, clip]{GeneralSources/LastFrameDockingTwoRingsFiveMembers.jpg}
	\caption{Structural ilustration of the molecule identied as \ref{figureCelecoxibTwoRings}.7 and \gls{cox2} docking generated by the ChimeraX software[15].}
	\label{figureDockingBestResult}
\end{figure}

\newpage
\section{Conclusions}

@@

@@

\begin{table}[H]
    \centering	
	\caption{Conditions and statistics for the computation of the results from tables (\ref{TableMainResults1}) and (\ref{TableMainResults2}).}
	\label{TableConditionsMainResults}
	\begin{tabular}{c|c}
		NumberOfTrees            & 250\\\hline
		Erased Percentatge       & 0.0\%\\\hline
		Splitting proportion     & 10.0\% for testing \\\hline
		minimumCorrelationFactor & 0.0\\\hline
		Number of descriptors    & 3040\\\hline
		Mean Squared Error       & 1202.70\\\hline
		$r^2$	                 & 0.90\\\hline
		True Positive            & 994\\\hline
		False Positive           & 8\\\hline
		True Negative            & 1015\\\hline
		False Negative           & 101\\\hline
		True Positive Rate       & 0.91\\\hline
		True Negative Rate       & 0.99\\\hline
		ClassificationAcuracy    & 0.94\\\hline
		MatthewsCorrelationFactor& 0.87
    \end{tabular}
\end{table}

>>>>

%-------------LINKS AND LOOKS-----------%
\usepackage{fullpage}
\usepackage{afterpage}
\usepackage{lastpage}
\usepackage{fancyhdr}
\usepackage{forest}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage{hyperref}
\usepackage{pgfplots}
\usepackage{xurl}
\usepackage{pdflscape}
\usepackage{overpic}
%-----------------FIGURES AND TABLES----------------%
\usepackage{array}
%\usepackage{wrapfig}
\usepackage{caption}
\DeclareCaptionType{scheme}
%\usepackage{subcaption}
\usepackage{tabularx, booktabs, makecell, multicol, multirow, bigstrut} %for excel2latex
\usepackage{float}

@@

Determining an experimental \gls{ic50} value can be both costly and time-consuming\footnote{Usually this parameter is computed throughout the Cheng Prusoff Equation (Def. (\ref{definitionChengPrusoff})) with experimental data.}. On the other hand, \gls{ai} provides an alternative by offering highly accurate predictions based on existing data, optimising research processes, and accelerating scientific discovery. In this scenario this project aims to implement artificial intelligence in computational chemistry, concretely, using \gls{ai}-based algorithms to predict a drug’s inhibition potential\cite{BookIC50} for a given protein with a relativelly good accuracy \footnote{The accuracy of the ML models is discussed on the Results and Discussion section (\ref{sectionResultsAndDiscussion}).}. To achieve this, we make use of the ChEMBL database\cite{ChemblDatabase}, a vast repository of bioactive molecules with drug-like properties. We extract all known molecular data with a documented inhibition potential for the target protein, creating a comprehensive dataset. The chemical descriptors (Def. (\ref{definitionChemicalDescriptors})) of each molecule in the database are then computed using AlvaDesk\cite{MauriMolecularDescriptorsBook}\cite{AlvaDescSecondPaper} software. Around 5000 descriptors are calculated\cite{DescriptorsBook}, which comprehend from the elemental molecular weight to the complex equipotential electronic surface, providing critical information about each compound’s behaviour. The resulting dataset is subsequently used to train \gls{ai} models, enabling them to predict the inhibition potential of unknown compounds. Finally, we evaluate the reliability of each model by testing it against real experimental data.

@@

\begin{figure}[H]
	\centering
		\begin{overpic}[width = \textwidth, trim={0.85cm 0.3cm 1.5cm 1.35cm}, clip]{../Plots/DataGapRepresentationIC50percentageEresed15.0Gap_min_32Gap_max_53.pdf}
		\put(10,55){Highly active}
		\put(30,55){Gap}
		\put(55,55){Least active}
	\end{overpic}
	\caption{\gls{ic50} values from our dataset and generation of a 20\% gap.}
	\label{figureGraphicRecurrecyIC50}
\end{figure}

@@

\begin{figure}[H]
\begin{tikzpicture}
\begin{axis}[
    width=0.45\textwidth,
    height=8cm,
    xlabel={$\%_{\text{Erased}}$},
    ylabel={$r^2$},
    xmin=0, xmax=100,
    ymin=0.75, ymax=1.0,
    xtick={0, 10, 20, 30, 40, 50, 60, 70, 80, 90, 100},
    legend pos=south east,
    legend style={font=\small},
    tick label style={font=\small},
    label style={font=\small},
]
\addplot[
    color=magenta,
    mark=diamond*,
    thick
] table[x index=0, y index=1, col sep=space] {GeneralSources/ErasedPercentageVSRSquare.txt};
\end{axis}
\end{tikzpicture}
\begin{tikzpicture}
\begin{axis}[
    width=0.45\textwidth,
    height=8cm,
    xlabel={$\%_{\text{Erased}}$},
	ylabel={Mean Squared Error (nM$^2$)},
    xmin=0, xmax=100,
    ymin=0, ymax=200,
    xtick={0, 10, 20, 30, 40, 50, 60, 70, 80, 90, 100},
    legend pos=north east,
    legend style={font=\small},
    tick label style={font=\small},
    label style={font=\small}
]
\addplot[
    color=magenta,
    mark=diamond*,
    thick
] table[x index=0, y index=1, col sep=space] {GeneralSources/ErasedPercentageVSMeanSquaredError.txt};
\end{axis}
\end{tikzpicture}
\caption{Plots of $\%_{\text{Erased}}$ vs $r^2$ and $\%_{\text{Erased}}$ vs Mean Squared Error. For a maximum $IC_{50}$ = 200 nM and a minimum of 0 nM,  taking into account all the chemical descriptors, the 20.0\% of the remaining data is saved for testing and the  80.0\% for training. The training is done with 200 trees.}
\label{FigPlotErasedPercentageVSRSquare}
\end{figure}

@@

Furthermore, we may also show the statistics and reliability of this results in Table (\ref{TableStatisitcsMainResults}). In the statistics we can see how all values surpass 0.9. In particular the model is capable of predicting when a molecules is not a highly active inhibitor in the 99\% of the times, since the True Negative Rate (Def.(\ref{definitionTrueNegativeRate})) is 0.99. The classification accuracy (Def. \ref{definitionClassificationAccuracy}) and the Matthews Correlation Factors take values close to 1, remarking thes particular model as a quite reliabile \gls{rf} model.

\begin{table}[H]
    \centering	
	\caption{Statistics for the computation of the results from tables (\ref{TableMainResults1}) and (\ref{TableMainResults2}).}
	\label{TableStatisitcsMainResults}
	\begin{tabular}{c|c}
		Mean Squared Error       & 1202.70\\\hline
		$r^2$ 	                 & 0.90\\\hline
		True Positive            & 994\\\hline
		False Positive           & 8\\\hline
		True Negative            & 1015\\\hline
		False Negative           & 101\\\hline
		True Positive Rate       & 0.90\\\hline
		True Negative Rate       & 0.99\\\hline
		ClassificationAcuracy    & 0.95\\\hline
		MatthewsCorrelationFactor& 0.90
    \end{tabular}
\end{table}

At this stage, we must point out the model is unable to distinguish between the active and unactive configurations. This problem is not related cased by the model or the descriptors computation but the database the model is trained with. The lack of existence of -cis/-trans configurations in the ChEMBL database \cite{ChemblDatabase} makes it impossible for the model to know the -cis conformation to be different from the -trans conformation. In other words, the model is unable to generate information, if it is not told to distinguish between -cis and -trans configurations, then it will not distinguish between them. Therefore the model will also provide the \gls{ic50} predictions for the active configuration, i.e. the -cis configuration.
Our hypothesis holds, since in Figure (\ref{figureGraphicGbindingIC50}) we can see a clearly a correlation between the \gls{ic50} and $\Delta G_{binding}$ matching with our initial intuition. Lower values of free eneregy (the most negative ones) correspond to low \gls{ic50} values i.e. highly active molecules.

However, the hipothesis of the existence of a combination of chemical descriptors that is directly related with a molecule's inhibition power for a given protein seems correct. Although, finding this combination of chemical descriptors is not an easy task for a chemist. At first glace, we might hypothesize that since this combination must be encoded into the \gls{rf} models unencripting this models would help us to understand the insights of the \gls{cox2}\footnote{Or any other protein since the code is through to be applied to any protein with an entrance in the ChEMBL database\cite{ChemblDatabase}.}.

Still this procedure is almost impossible, since the nature of the \gls{rf} is similar to check weather a pasword is correct or not. In other words, computationally speaking, predicting a molecules \gls{ic50} as well as checking an username and a pasword is an easy task for the computer. Nevertheless, reathriving a password and highlight the chemical descriptors with higher dependecy on the \gls{ic50} is another story.

For instance, one could think making predictions of random molecules \gls{ic50} by varying one molecular descriptor and holding the resest constant could give us information on this dependecy\footnote{This would be a desencriptation by brute force.}. However, we encounter several unsortable incoherences in this procedure. For example, let us think we want to observe the dependecy of the \gls{ic50} on a molecules' average bond distance. Molecules that, at first, would not fit into the binding cavity of the \gls{cox2} could happend to fit after this streaching, but this molecules are not real, i.e. non-stable, lossing any chemical coherence. Additionaly there are some chemical descriptors that integers by nature, for istance the number of carbons in a molecule, but this procedure would point out the ideal number of carbons to be a good inhibitor is 25.5 carbons, and absurd outcome. Furthermore a set arbitrarly generated chemical descriptors it is not necessarily correlated to a molecule, in the majority of the cases there will not be any molecule to have the specified chemical descriptors, thus we would be trying to predict the \gls{ic50} of imaginary molecules.

It is important to remark this observations does not deny the power of \gls{ai} for computational chemistry, but points out at some limitations of the technology.

@@

After the predictions have been computed a docking has been performed of one of the most promising azophotoswitches. In particular the docking of the  \ref{figureCelecoxibTwoRings}.7 molecule depicted in Scheme (\ref{figureCelecoxibTwoRingsDoking}), this molecule has a computed $\Delta G_{binding}$ of 10.6712 Kcal/mol and a predicted \gls{ic50} of 77 nM. An image of the last frame of the docking procedure is provided in Figure (\ref{figureDockingBestResult}).
\begin{figure}[H]
\captionsetup{type=scheme}
\centering
\scalebox{0.7}{
\chemfig[angle increment=30]{
	H_2N-S(=[3]O)(=[9]O)-*6(-=-(-N=[-1.5]N-[-3]*5(=-O-*5(-N(-[-3]H)-(-)=(-*6(-=-(-)=-=))-=)--))=-=)
}}
\caption{Template for Celecoxib azo-derivatives highly active molecule.}
\label{figureCelecoxibTwoRingsDoking}
\end{figure}

\begin{figure}[H]
	\centering
	\includegraphics[width = \textwidth, trim={0cm 0cm 0cm 0cm}, clip]{GeneralSources/LastFrameDockingTwoRingsFiveMembers.jpg}
	\caption{Structural ilustration of the molecule identied as \ref{figureCelecoxibTwoRings}.7 and \gls{cox2} docking generated by the ChimeraX software[15].}
	\label{figureDockingBestResult}
\end{figure}
Hence we can see the fitting of the azophotoswitch in the -cis configuration inside the \gls{cox2} cavity. Although the -trans configuration has been seen to fit into the binding site\footnote{With few problems} it is impossible for this isomer to access to the binding site due to stereochemical problems, i.e. the -trans conformation colides with the outer layers of the \gls{cox2} and cannot access to the binding cavity. Still this molecule could interact with undesired proteins in both configurations (E)/(Z) which could lead to enormous side effects if it is administered as a drug, then further investigations must be done.

\newpage
\section{Conclusions}
We have proved Artificial Intelligence's potential for computational and theorical chemistry. Furthermore, we have developed a totally functional Machine Learning software capable of predicting \gls{ic50} values for any protein in the ChEMBL database. Also the software is capable of generating the statistics for each \gls{ml} model generated defined by the hard-coded variables that define the data the model is trained with. Thus the user is capable of playing arround with the variables in order to obtain the best predictions possible.

Additionaly we have proved our intuition on the \gls{ic50} direct relationship with the provided $\Delta G_{binding}$ values to be correct. Furthermore, a docking of one of the most promising azophotoswitches as been performed matching our expectations and reinforcing the \gls{ai}'s potential.

@@

\begin{definition}\label{definitionChemicalDescriptors}
Molecular descriptor: "A molecular descriptor is the final result of a logical and mathematical procedure which transforms chemical information encoded within a symbolic representation of a molecule into a useful number or the result of some standardized experiment."\cite{DescriptorsBook}
\end{definition}

@@

\begin{table}[H]
    \centering	
	\caption{Conditions and statistics for the computation of the results from tables (\ref{TableMainResults1}) and (\ref{TableMainResults2}).}
	\label{TableConditionsMainResults}
	\begin{tabular}{c|c}
		NumberOfTrees            & 250\\\hline
		Erased Percentatge       & 0.0\%\\\hline
		Splitting proportion     & 10.0\% for testing \\\hline
		minimumCorrelationFactor & 0.0\\\hline
		Number of descriptors    & 3040\\\hline
		Mean Squared Error       & 1202.70\\\hline
		$r^2$ 	                 & 0.90\\\hline
		True Positive            & 994\\\hline
		False Positive           & 8\\\hline
		True Negative            & 1015\\\hline
		False Negative           & 101\\\hline
		True Positive Rate       & 0.90\\\hline
		True Negative Rate       & 0.99\\\hline
		ClassificationAcuracy    & 0.95\\\hline
		MatthewsCorrelationFactor& 0.90
    \end{tabular}
\end{table}
